---
title: "The Art and Science of p-value Hacking"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
  | \faTwitter\ ```@xmjandrews```
  | \faGithub\ ```https://github.com/lawsofthought/smlp2017```
date: "August 31, 2017"
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: slides_preamble.tex
bibliography: refs.bib
csl: apa.csl
---


```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = FALSE, include=FALSE)
```

## P-hacking: A definition

- *p-hacking*, broadly defined, is the manipulation, whether intentional or not, of frequentist[^1] statistical testing procedures in order to obtain a desired outcome.
- Technically, p-hacking is always a type of undisclosed multiple simultaneous statistical testing, whereby the de facto false positive rate greatly exceeds the nominal false-positive rate, i.e. $\alpha$.
- Related terms include *data-dredging*, *data-snooping*, *data-fishing*, *cherry-picking*, *significance-chasing*, and so on.
- The original contemporary exposé of this general phenomenon is due to @simmons2011false, followed by @simonsohn2014p.


[^1]: The Bayesian counterpart to p-hacking is b-hacking.


## Frequentist statistical testing
\framesubtitle{Neyman-Pearson testing}

- In advance of data collection, our scientific hypothesis, operationalized as $\mathcal{H}_1$ (e.g. $\mathcal{H}_1: \theta > 0$), is specified. 
- This leads to a corresponding null hypothesis, e.g. $\mathcal{H}_0: \theta = 0$, a test statistic $T(\mathcal{D})$, and a critical threshold for the statistic $T_{\text{crit}}$, such that
$$
\Prob{\vert T(\mathcal{D}) \vert > T_{\text{crit}} \given \mathcal{H}_0 = \text{True}} = \alpha,
$$
where $\alpha$ is conventionally $0.05$.
- We then determine a minimum sample size for $\mathcal{D}$ so that 
$$
\Prob{\vert T(\mathcal{D}) \vert > T_{\text{crit}} \given \mathcal{H}_1 = \text{True}} \gtrapprox 1 - \beta,
$$
where $\beta$ is conventionally $0.2$ or $0.1$.
- We then collect $\mathcal{D}$, calculate $T(\mathcal{D})$. If $\vert T(\mathcal{D}) \vert > T_{\text{crit}}$ then we reject the null. Otherwise, we do not reject it.
- Following this procedure, in the long run, our false positive rate will be $\alpha$, and our false negative rate will be $\beta$.



## P-hacking

- In advance of data collection, we begin with a (perhaps vaguely stated) scientific hypothesis.
- We collect data $\mathcal{D}$.
- We then operationalize our scientific hypothesis as $\mathcal{H}^{k}_1$, which leads to $\mathcal{H}^{k}_0$, $T^{k}(\mathcal{D})$, $T^{k}_{\text{crit}}$, starting with $k=1$.
- We calculate $T^k(\mathcal{D})$. If $\vert T^k(\mathcal{D}) \vert > T^k_{\text{crit}}$ then we reject the null and stop. 
- Otherwise, if $\vert T^k(\mathcal{D}) \vert \leq T^k_{\text{crit}}$, we re-operationalize our scientific hypothesis as $\mathcal{H}^{k=2}_1$ and test again.
- We continue as such indefinitely and stop when we obtain a significant result, and then report *only* that result.
- Following this procedure, in the long run, our false positive rate will be $\gg \alpha$.


## P-hacking example 1: Subsetting

```{r, cache=TRUE}
set.seed(10001)
alpha <- 0.05
iterations <- 25000
n <- 20
is.significant <- function(M) summary(M)$coefficients['independent.var','Pr(>|t|)'] < alpha

any.significant.subsetting <- function(n){
  
  dependent.var <- rnorm(2*n)
  
  independent.var <- c(rep(0, n),
                       rep(1, n))
  
  gender <- round(runif(2*n))
  
  Df <- data.frame(dependent.var, 
                   independent.var, 
                   gender)
  
  M.0 <- lm(dependent.var ~ independent.var, data = filter(Df, gender==1))
  M.1 <- lm(dependent.var ~ independent.var, data = filter(Df, gender==0))
  M.2 <- lm(dependent.var ~ independent.var, data = Df)
  
  any(sapply(list(M.0, M.1, M.2), is.significant))
  
}

false.positive.rate.subset <- mean(replicate(iterations, any.significant.subsetting(n)))

```

Online demo: https://lawsofthought.shinyapps.io/p_hacking/ 

- Let's assume we want to test if two groups of people differ in the mean value of some variable.
- In both groups, there are men and women. 
- We can test just the men, just the women, or both.
- In a simulation with $n=`r n`$ people in each group, with $\alpha=0.05$, subsetting results in the false positive rate being $\approx$ `r round(false.positive.rate.subset*100, 1)`%.

## P-hacking example 2: Adding a covariate

```{r, cache=TRUE}
set.seed(1001)
alpha <- 0.05
n <- 20
iterations <- 25000

covariate.phack <- function(n){
  
  main.effect.significant <- function(M) summary(M)$coefficients['independent.var','Pr(>|t|)'] < alpha
  interaction.significant <- function(M) summary(M)$coefficients['independent.var:gender','Pr(>|t|)'] < alpha
  
  dependent.var <- rnorm(2*n)
  
  independent.var <- c(rep(0, n),
                       rep(1, n))
  
  gender <- round(runif(2*n))
  
  M.0 <- lm(dependent.var ~ independent.var)
  M.1 <- lm(dependent.var ~ independent.var + gender)
  M.2 <- lm(dependent.var ~ independent.var * gender)
  
  any(c(sapply(list(M.0, M.1, M.2),
               main.effect.significant),
        interaction.significant(M.2)))
  
}

covariate.phack.pcurve <- function(n){
  
  main.effect.pvalue <- function(M) summary(M)$coefficients['independent.var','Pr(>|t|)']
  interaction.pvalue <- function(M) summary(M)$coefficients['independent.var:gender','Pr(>|t|)']
  
  dependent.var <- rnorm(2*n)
  
  independent.var <- c(rep(0, n),
                       rep(1, n))
  
  gender <- round(runif(2*n))
  
  M.0 <- lm(dependent.var ~ independent.var)
  M.1 <- lm(dependent.var ~ independent.var + gender)
  M.2 <- lm(dependent.var ~ independent.var * gender)
  
  p.values <- c(sapply(list(M.0, M.1, M.2),
           main.effect.pvalue),
    interaction.pvalue(M.2))
  
  if (any(p.values < alpha)){
    return(p.values[p.values < alpha][1]) # Return first one
  } else {
    return(NULL)
  }
  
}


p.curve.values <- unlist(replicate(iterations, covariate.phack.pcurve(n)), recursive = T)
false.positive.rate.covariate <- mean(replicate(iterations, covariate.phack(n)))

```

- In an identical problem to before, instead of subsetting by gender, we simply add it as a covariate.
- We then test if the main effect exists in the presence and absence of the gender covariate, or if there is an interaction between gender and the main effect. 
- In another simulation, $n=`r n`$ people in each group, with $\alpha=0.05$, this leads to the false positive rate being $\approx$ `r round(false.positive.rate.covariate*100, 1)`%. 

## P-hacking example 3: Optional stopping

Online demo: https://lawsofthought.shinyapps.io/optional_stopping/

```{r, include=TRUE,  out.width = "220px", fig.align="center"}

z.test <- function(x){
  n <- length(x)
  z <- mean(x) * sqrt(n)
  pnorm(abs(z), lower.tail=F) * 2.0
}

is.significant <- function(x, alpha){
  ifelse(z.test(x) <= alpha, TRUE, FALSE)
}

sample.until.significance <- function(initial.sample.size = 10,
                                      alpha=0.05,
                                      max.sample.size = 100,
                                      sample.incrementation=10){

  # Draw initial sample
  x <- rnorm(initial.sample.size)

  # Is the initial sample significant?
  significant <- is.significant(x, alpha)

  # Repeat until significant (or maximum sample size exceeded)
  while (!significant & length(x) < max.sample.size){

    x <- c(x, rnorm(sample.incrementation))
    significant <- is.significant(x, alpha)
  }

  return(significant)

}


error.rates.with.optional.stopping <- function(max.sample.size = 100,
                                               initial.sample.size = 10,
                                               sample.size.increment = 10,
                                               alpha = 0.05,
                                               iterations = 10000){

  I <- seq(initial.sample.size, max.sample.size, by=sample.size.increment)

  error.rates <- apply(
    replicate(iterations, {
      # Draw max-size sample first
      x <- rnorm(max.sample.size)

      # Calculate the Z statistic after each additional data point
      # This is (the abs value of) the cumulative mean times the
      # sqrt of the cumulative sample size
      z.stats <- abs(cummean(x) * sqrt(seq(length(x))))

      # Calculate the significance after each addtional data point
      significant <- z.stats >= qnorm(1-alpha/2.0)

      # Examine significance at initial.sample.size and every sample.size.increment
      # steps until we reach max.sample.size
      s <- significant[I]

      # A cumulative sum at each value of s tells us how many
      # significant values we've seen until that point.
      # If there is more than 0, then that means we would have
      # stopped at declared a significant result at that stage.
      cumsum(significant[I]) > 0
    }), 1, mean)

  data.frame(sample.size = I,
             error.rate = error.rates)

}

s <- error.rates.with.optional.stopping(initial.sample.size = 10,
                                        max.sample.size = 100,
                                        sample.size.increment = 5,
                                        alpha = 0.05)

g0 <- ggplot(s, mapping=aes(x = sample.size, y=error.rate)) + 
  theme_classic() + 
  xlab('Sample size') + 
  ylab('Probability of significant result')

g0 + geom_bar(stat='identity')

```

- Collecting data, testing, and then collecting more data if results are not significant, leads to a steady rise in false positive rates.

## P-hacking example 4: Removing outliers

```{r, cache=TRUE, include=TRUE}
set.seed(12345)
n <- 20
iterations <- 25000
alpha <- 0.05
  
sd.outliers <- function(x, k=2.0){
  (x > (mean(x) + k*sd(x))) | (x < (mean(x) - k*sd(x)))
}

quantile.outliers <- function(x, delta=0.05){
  (x > quantile(x, 1-delta)) | (x < quantile(x, delta))
}

k.outliers <- function(x, k=5){
  (rank(x) > (length(x)-k)) | (rank(x) < k )
}
  
any.significant <- function(n){
  
  dependent.var <- rnorm(2*n)
  independent.var <- c(rep(0, n),
                       rep(1, n))
  
  Df <- data.frame(dependent.var,
                   independent.var)
  
  Df.0 <- Df
  Df.1 <- Df[!sd.outliers(Df$dependent.var, k=2.0),]
  Df.2 <- Df[!sd.outliers(Df$dependent.var, k=1.5),]
  Df.3 <- Df[!quantile.outliers(Df$dependent.var, delta=0.025),]
  Df.4 <- Df[!quantile.outliers(Df$dependent.var, delta=0.05),]
  Df.5 <- Df[!k.outliers(Df$dependent.var, k=5),]
  Df.6 <- Df[!k.outliers(Df$dependent.var, k=2),]
  
  is.significant <- function(Df) summary(lm(dependent.var ~ independent.var, data=Df))$coefficients['independent.var','Pr(>|t|)'] < alpha
  
  any(sapply(list(Df.0, Df.1, Df.2, Df.3, Df.4, Df.5, Df.6),
             is.significant))
  
}

false.positive.rate.outliers <- mean(replicate(iterations, any.significant(n)))
```

- Using an identical problem to before, we remove outliers, or not, before testing.
- Outlier may be defined as any of the following:
    1. Data above/below 2 SDs from mean.
    2. Data above/below 1.5 SDs from mean.
    3. Data in the upper/lower 5% quantiles.
    4. Data in the upper/lower 5% quantiles.
    5. The 2 highest/lowest values.
    6. The 5 highest/lowest values
- In a simulation, with $n=`r n`$ in each group, and $\alpha = `r alpha`$, this leads to a false positive rate of `r round(100*false.positive.rate.outliers, 1)`%. 


## P hacking broadside 
\framesubtitle{Combine your p-hack tools for maximum effect}

```{r, cache=TRUE, include=TRUE}
set.seed(101101)
alpha <- 0.05
n <- 20
k <- 10
N <- 100
iterations <- 250

gen.data <- function(n){
  data.frame(dependent.var = rnorm(2*n),
             independent.var = c(rep(0, n),
                                 rep(1, n)),
             gender = round(runif(2*n)))
}

covariate.phack <- function(Df){
  
  main.effect.significant <- function(M) summary(M)$coefficients['independent.var','Pr(>|t|)'] < alpha
  interaction.significant <- function(M) summary(M)$coefficients['independent.var:gender','Pr(>|t|)'] < alpha
  
  M.0 <- lm(dependent.var ~ independent.var, data=Df)
  M.1 <- lm(dependent.var ~ independent.var + gender, data=Df)
  M.2 <- lm(dependent.var ~ independent.var * gender, data=Df)
  
  any(c(sapply(list(M.0, M.1, M.2),
               main.effect.significant),
        interaction.significant(M.2)))
  
}

drop.outliers <- function(Df){
  
  Df.0 <- Df
  Df.1 <- Df[!sd.outliers(Df$dependent.var, k=2.0),]
  Df.2 <- Df[!sd.outliers(Df$dependent.var, k=1.5),]
  Df.3 <- Df[!quantile.outliers(Df$dependent.var, delta=0.025),]
  Df.4 <- Df[!quantile.outliers(Df$dependent.var, delta=0.05),]
  Df.5 <- Df[!k.outliers(Df$dependent.var, k=5),]
  Df.6 <- Df[!k.outliers(Df$dependent.var, k=2),]
  
  list(Df.0, Df.1, Df.2, Df.3, Df.4, Df.5, Df.6)
}

outlier.covariate.phack <- function(n){
  
  Df <- gen.data(n)
  any(sapply(drop.outliers(Df), covariate.phack))
  
}

outlier.covariate.phack.optional.stopping <- function(n, k, N){
  
  Df <- gen.data(n)
  significant <- any(sapply(drop.outliers(Df), covariate.phack))
  
  while (!significant & dim(Df)[1] < 2*N){
    Df <- rbind(Df, gen.data(k))
    significant <- any(sapply(drop.outliers(Df), covariate.phack))
  }
  
  significant
  
}

double.cocktail <- mean(replicate(iterations, outlier.covariate.phack(n)))
triple.cocktail <- mean(replicate(iterations, outlier.covariate.phack.optional.stopping(n, k, N)))

```

- Using an identical problem to before, we start with two samples of size $n=`r n`$ and $\alpha = `r alpha`$.
- Combining our removal of outliers method *and* our covariate method leads to a false positive rate of `r round(100*double.cocktail, 1)`%.
- Combining our removal of outliers method *and* our covariate method *and* collecting `r k` new data points in each group until significance or `r N` in each group leads to a false positive rate of `r round(100*triple.cocktail, 1)`%.

## P-value distribution (p-curves) under null

- The distribution of p-values (p-curves) in any given body of work will be a function of whether the true effect size, which may be zero, and the extent of p-hacking.
- Whether we can use p-curves in meta-analysis to assess the extent and consequences of p-hacking, as recommended by @simonsohn2014p, is a matter of debate, see 
    - @gelman2013discussion
    - @head2015extent
    - @bishop2016problems
    - @bruns2016p
    - @hartgerink2017reanalyzing

## P-curve under null

```{r, cache=TRUE}
iterations <- 1e5

initial.sample.size <- 10
alpha <- 0.05
max.sample.size <- 100
sample.incrementation <- 10

sample.until.significance.pvalue <- function(initial.sample.size = 10,
                                             alpha=0.05,
                                             max.sample.size = 100,
                                             mu = 0.0,
                                             sample.incrementation=10){
  
  # Draw initial sample
  x <- rnorm(initial.sample.size, mean=mu)
  
  # Is the initial sample significant?
  significant <- is.significant(x, alpha)
  
  # Repeat until significant (or maximum sample size exceeded)
  while (!significant & length(x) < max.sample.size){
    
    x <- c(x, rnorm(sample.incrementation, mean=mu))
    p.value <- z.test(x)
    significant <- ifelse(p.value < alpha, TRUE, FALSE)
  }
  
  return(z.test(x))
  
}

non.phacked.pvalues <- replicate(iterations,
                                 z.test(rnorm(initial.sample.size))
)

non.phacked.pvalues.nonnull.med <- replicate(iterations,
                                             z.test(rnorm(initial.sample.size, mean=0.5))
)

non.phacked.pvalues.nonnull.low <- replicate(iterations,
                                             z.test(rnorm(initial.sample.size, mean=0.2))
)

phacked.pvalues.null <- replicate(iterations, 
                                  sample.until.significance.pvalue(initial.sample.size,
                                                                   alpha,
                                                                   max.sample.size,
                                                                   mu = 0.0,
                                                                   sample.incrementation)
)

phacked.pvalues.nonnull.low <- replicate(iterations, 
                                         sample.until.significance.pvalue(initial.sample.size,
                                                                          alpha,
                                                                          max.sample.size,
                                                                          mu = 0.2,
                                                                          sample.incrementation)
)

phacked.pvalues.nonnull.med <- replicate(iterations, 
                                         sample.until.significance.pvalue(initial.sample.size,
                                                                          alpha,
                                                                          max.sample.size,
                                                                          mu = 0.5,
                                                                          sample.incrementation)
)

plot.pcurve <- function(pvals){
  Df <- data.frame(p.values = pvals)
  dplyr::filter(Df, p.values <= alpha) %>% 
    ggplot(mapping = aes(x=p.values)) + 
    geom_histogram(breaks = seq(0.0, 0.05, by=0.003)) + 
    theme_classic()
}


```
When null is true, the distribution of p-values is uniform.
```{r, include=TRUE, echo=FALSE, message=FALSE}
plot.pcurve(non.phacked.pvalues)
```

## P-value distribution (p-curve) under non-null
\framesubtitle{Medium effect}

When null is false, the distribution of p-values is right skewed.
```{r, include=TRUE, echo=FALSE, message=FALSE}
plot.pcurve(non.phacked.pvalues.nonnull.med)
```

## P-curve under non-null
\framesubtitle{Low effect}

When null is false, the distribution of p-values is right skewed.
```{r, include=TRUE, echo=FALSE, message=FALSE}
plot.pcurve(non.phacked.pvalues.nonnull.low)
```

## P-curve under null with p-hacking

When null is true, the distribution of \emph{p-hacked} p-values is left-skewed.
```{r, include=TRUE, echo=FALSE, message=FALSE}
plot.pcurve(phacked.pvalues.null)
```

## P-curve under non-null with p-hacking
\framesubtitle{Low effect}

When effect is low, the distribution of \emph{p-hacked} p-values is left-skewed.
```{r, include=TRUE, echo=FALSE, message=FALSE}
plot.pcurve(phacked.pvalues.nonnull.low)
```

## P-curve under non-null with p-hacking
\framesubtitle{Medium effect}

When effect is medium, the distribution of \emph{p-hacked} p-values is right-skewed.
```{r, include=TRUE, echo=FALSE, message=FALSE}
plot.pcurve(phacked.pvalues.nonnull.med)
```

## Prevalence of P-hacking

\begin{figure}[placeHere]
  \centering
    \includegraphics[width=\textwidth]{"figs/john_loewenstein_prelec"}
\end{figure}
From @john2012measuring.

## Why P-hacking is so toxic for science

- P-hacking is easy to do (you just need some ethical laxity and some stamina)
- It is hard to detect 
- It can dramatically increase the false positive rate
- False positives are hard to detect and hard to eliminate 
- False positives add noise to the literature, and result in wasted resources when used as the basis for future research
- P-hacking may be self-perpetuating: Results are p-hacked because some effects are assumed to be real (on the basis of p-hacked literature)

## How to eliminate p-hacking?

- P-hacking is an ethical problem, rather than a statistical issue. 
- P-hacking can be eliminated by changing ethical standards:
    - *Honesty in reporting*: The explicit recommendations in @simmons2011false are largely recommendations for a cultural shift away from selective reporting.
    - *Pre-registration*: It immediately eliminates *harking* and greatly reduces researcher degrees of freedom
    - *Open (raw) data and analysis code*: Disclosing all the original data (especially as recommended by @rouder2016and) and the processing/analysis pipeline can make tricks easier to identify, and allows alternative analyses to be performed

## References {.allowframebreaks}
